Revisiting the hydrodynamic modulation of short surface waves by longer waves

### 2025-03-06 paper.tex
Text revisions

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the linear wave theory and derive higher-order
steady solutions for the change of short-wave wavenumber, action, and
gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed to be 5/8 due to the
wavenumber, 1/4 due to the wave action, and 1/8 due to the effective gravity.
We further examine the result of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx$0.2.
Except for a subset of initial conditions, the solutions to the fully nonlinear
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_hydrodynamic_modulation_diagram.pdf}
\caption{
  A diagram of short waves riding on longer waves that propagate with their
  phase-speed $C_{pL}$ from left to right.
  The short waves propagate on the long-wave surface, their action
  moving with the group velocity $C_g$.
  Long waves induce orbital velocities $U$ that cause near-surface convergence
  and divergence on their front and rear faces, respectively.
  They also induce downward and upward centripetal accelerations
  ($\partial W/\partial t$) in the crests and troughs, respectively, that
  modulate the effective gravity at the surface.
  As a consequence of surface convergence and divergence and the effective-gravity
  modulation, the short waves become shorter and higher preferentially
  on the long-wave crests.
  The diagram is not to scale.
}
\label{fig:hydrodynamic_modulation_diagram}
\end{figure}

All modulation solutions mentioned thus far are steady in the reference frame
of the long wave---short waves become shorter and higher (and thus, steeper)
preferentially on the long-wave crests.
Conversely, they elongate and flatten preferentially in the long-wave troughs.
Recently, \citet{peureux2021unsteady} asked whether the steady solutions
of short-wave modulation are appropriate by simulating the full wave conservation
equations.
They found that the short waves grow unsteadily due to the propagation of
longer waves, suggesting that the steady-state solutions may only be valid for
a few long-wave periods before the short waves begin to break due to excessive
steepness.
However, the unsteadiness of their solutions occurs in a specific scenario in
which a long-wave train suddenly appears to modulate a uniform field of short
waves.
The numerical simulations stabilise if the short-wave field is initialised
from existing steady solutions, like those by \citet{longuet1960changes}.
Nevertheless, their results do put into question whether the short-wave
modulation predicted by theory is generally steady over many long-wave periods
or not.
Looking at the conservation equations alone, it is not obvious that it is.

@@

\begin{equation}
\label{eq:k_short_lhs}
\frac{\widetilde{k}}{k} = 1 + \varepsilon_L \cos{\psi}
\end{equation}
Coming from the conservation of crests, as opposed to the velocity potential
superposition, the solution by \citet{longuet1960changes} requires two
approximations:
First, evaluating the long-wave velocity at $z = 0$ rather than $z = \eta_L$; and
second, further truncating the Taylor expansion series beyond the first-order
term.
These approximations underestimate the short-wave modulation magnitude
(Fig. \ref{fig:analytical_solutions_ak0.1}a).
At the wave crest, (\ref{eq:k_short_exact}) yields the wavenumber modulation
that is 16.9\%, 38.5\%, 66.4\%, and 104\% larger than that of
\citet{longuet1960changes} for $\varepsilon_L = 0.1, 0.2, 0.3, 0.4$, respectively.
Although not immediately obvious from the expressions, the modulated wavenumber
(\ref{eq:k_short_lhs}) is conserved across the long-wave phase, whereas
the higher-order solution (\ref{eq:k_short_exact}) is not.
This is because combining the orbital velocities evaluated at $z=\eta_L$ and the
linearised conservation of wave crests in (\ref{eq:wavenumber}) creates excess
short-wave wavenumber.
The solution of (\ref{eq:wavenumber}) must be conservative across the long-wave
phase, as we will see later from the numerical simulations.

@@

Let us now examine the contributions of the different terms to the short-wave
effective gravity modulation.
Fig. \ref{fig:effective_gravities}a shows the effective gravity modulation of
short waves as a function of the long-wave phase for $\varepsilon_L = 0.2$, for
a variety of the gravity modulation forms.
This value of $\varepsilon_L$ allows easier visualisation of the modulation
differences, while still being relatively moderate.
To the first order in $\varepsilon_L$, the gravity modulation is out of phase
with the long-wave elevation, with the short waves experiencing a lower effective
gravity on the long-wave crests and a higher effective gravity in the troughs.
Evaluating the orbital accelerations at $z=\eta_L$ causes a gravity reduction
around the crests ($\approx6\%$) and the troughs ($\approx 3\%$), and an
increase on the front and rear faces ($\approx 4\%$) of the long wave, due to
the exponential change in the orbital accelerations away from the mean water level
(Fig. \ref{fig:effective_gravities}b).
The difference between the linear and Stokes long waves at this steepness is
negligible.
The long-wave elevation curvature has a second-order effect in $\varepsilon_L$.
Naturally, there is no curvature exactly at the crests and the troughs, and its
effects are largest at the front and rear faces of the long wave, where the
covariance between the local slope and the horizontal accelerations are largest.

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{figures/fig_effective_gravities.pdf}
\caption{
  (a) Analytical solutions for the effective gravitational acceleration
  modulation by long waves, as function of long-wave phase, for $\varepsilon_L = 0.2$.
  Black is for linear wave evaluated at $z=0$,
  blue and green are for linear and Stokes waves, respectively, evaluated at $z=\eta$ and without curvature effects,
  and orange and green are for linear and Stokes waves, respectively, evaluated at $z=\eta$ and with curvature effects.
  (b) As in (a) but normalised by the modulation of the linear wave evaluated at $z=0$, $g_1$.
}
\label{fig:effective_gravities}
\end{figure}

It is instructive to also examine the mean, maximum, and minimum values of the
effective gravity as function of long-wave steepness $\varepsilon_L$
(Fig. \ref{fig:effective_gravities_mean_max_min}).
The long-wave averaged effective gravity of short waves remains equal to the
resting value of $g$ for all values of $\varepsilon_L$, except if the effects
of curvature are considered, in which case the average effective gravity is
reduced.
This reduction, in case of a long Stokes wave, is $\approx 3\%$ for
$\varepsilon_L = 0.2$ and $\approx 14\%$ for $\varepsilon_L = 0.4$.
As the effective gravity minima and maxima occur at crests and troughs,
respectively, the curvature affects the phase-averaged values but not the extremes.
As the slope effects on the effective gravity exist only away from crests and
troughs, they may be neglected in the steady solutions where modulation is
typically evaluated at the long-wave crests. 
The effective gravity reduction at the crests is $\approx 11\%$, $25\%$, and $66\%$
for $\varepsilon_L = 0.1$, $0.2$, and $0.4$, respectively.
These are similar to the fully-nonlinear numerical estimates of
\citet{longuet1986eulerian,longuet1987propagation}.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/fig_effective_gravities_mean_max_min.pdf}
\caption{
  As Fig. \ref{fig:effective_gravities} but showing the mean (solid),
  maximum (dashed), and minimum (dotted) short-wave gravity modulation as
  function of long-wave steepness $\varepsilon_L$.
  The minimum and maximum values for the cases with curvature effects
  ($\alpha=\frac{\partial \eta}{\partial x}$) are equivalent to those without
  curvature effects ($\alpha=0$) and obscure them in the plot.
}
\label{fig:effective_gravities_mean_max_min}
\end{figure}

>>>>

\begin{abstract}
Hydrodynamic modulation of short ocean surface waves by longer ambient waves is
a well-known ocean surface process that affects remote sensing, the
interpretation of \textit{in situ} wave measurements, and numerical wave
forecasting.
In this paper, we revisit the wave crest and action conservation laws and derive
higher-order steady solutions for the change of short-wave wavenumber, action,
and gravitational acceleration due to the presence of longer waves.
We validate the analytical solutions with numerical simulations of the full wave
crest and action conservation equations.
The nonlinear analytical solutions of short-wave wavenumber, amplitude, and
steepness modulation significantly deviate from the linear analytical solutions
of \citet{longuet1960changes}, and are similar to the nonlinear numerical
solutions by \citet{longuet1987propagation} and \citet{zhang1990evolution}.
The short-wave steepness modulation is attributed to be 5/8 due to the
wavenumber, 1/4 due to the wave action, and 1/8 due to the effective gravity.
We further examine the results of \citet{peureux2021unsteady} who found through
numerical simulations that the short-wave modulation grows unsteadily with
each long-wave passage.
We show that this unsteady growth only occurs for homogeneous initial
conditions as a special case and does not generally occur, for example in more
realistic long-wave groups.
The proposed steady solutions are a good approximation of the fully
nonlinear numerical solutions in long-wave steepness up to $\approx$0.2.
Except for a subset of initial conditions, the solutions to the fully nonlinear
crest-action conservation equations are mostly steady in the reference frame of
the long waves.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_hydrodynamic_modulation_diagram.pdf}
\caption{
  A diagram of short waves riding on longer waves that propagate with their
  phase-speed $C_{pL}$ from left to right.
  The short waves propagate on the long-wave surface, their action
  moving with the group velocity $C_g$.
  Long waves induce orbital velocities $U$ that cause near-surface convergence
  and divergence on their front and rear faces, respectively.
  They also induce downward and upward centripetal accelerations
  ($dW/dt$) in the crests and troughs, respectively, that
  modulate the effective gravity at the surface.
  As a consequence of surface convergence and divergence and the effective-gravity
  modulation, the short waves become shorter and higher preferentially
  on the long-wave crests.
  The diagram is not to scale.
}
\label{fig:hydrodynamic_modulation_diagram}
\end{figure}

All modulation solutions mentioned thus far are steady in the reference frame
of the long wave---short waves become shorter and higher (and thus, steeper)
preferentially on the long-wave crests.
Conversely, they elongate and flatten preferentially in the long-wave troughs.
Recently, \citet{peureux2021unsteady} asked whether the steady (stationary) solutions
of short-wave modulation are appropriate by simulating the full wave conservation
equations.
They found that the short waves grow unsteadily due to the propagation of
longer waves, suggesting that the steady-state solutions may only be valid for
a few long-wave periods before the short waves begin to break due to excessive
steepness.
However, the unsteadiness of their solutions occurs in a specific scenario in
which a long-wave train suddenly appears to modulate a uniform field of short
waves.
The numerical simulations stabilise if the short-wave field is initialised
from existing steady solutions, like those by \citet{longuet1960changes}.
Nevertheless, their results do put into question whether the short-wave
modulation predicted by theory is generally steady over many long-wave periods
or not.
Looking at the conservation equations alone, it is not obvious that it is.

@@

\begin{equation}
\label{eq:k_short_lhs}
\frac{\widetilde{k}}{k} = 1 + \varepsilon_L \cos{\psi}
\end{equation}
Coming from the conservation of crests, as opposed to the velocity potential
superposition, the solution by \citet{longuet1960changes} requires two
approximations:
First, evaluating the long-wave velocity at $z = 0$ rather than $z = \eta_L$; and
second, further truncating the Taylor expansion series beyond the first-order
term.
These approximations underestimate the short-wave modulation magnitude
(Fig. \ref{fig:analytical_solutions_ak0.1}a).
At the wave crest, (\ref{eq:k_short_exact}) yields the wavenumber modulation
that is 16.9\%, 38.5\%, 66.4\%, and 104\% larger than that of
\citet{longuet1960changes} for $\varepsilon_L = 0.1, 0.2, 0.3, 0.4$, respectively.
Although not immediately obvious from the expressions, the modulated wavenumber
(\ref{eq:k_short_lhs}) is conserved across the long-wave phase, whereas
the higher-order solution (\ref{eq:k_short_exact}) is not.
This is because combining the orbital velocities evaluated at $z=\eta_L$ and the
linearised conservation of wave crests in (\ref{eq:wavenumber}) creates excess
short-wave wavenumber.
The solution of (\ref{eq:wavenumber}) must be conservative across the long-wave
phase, as we will see later from the numerical simulations.
Although the analytical solutions by \citet{longuet1960changes} has been updated
by the numerical simulations of \citet{longuet1987propagation}, they are here
nevertheless relevant to use as reference for the proposed analytical solutions.

@@

Let us now examine the contributions of the different terms to the short-wave
effective gravity modulation.
Fig. \ref{fig:effective_gravities}a shows the effective gravity modulation of
short waves as a function of the long-wave phase for $\varepsilon_L = 0.3$, for
a variety of the gravity modulation forms.
Although quite steep, this value of $\varepsilon_L$ allows easier visualisation
of the modulation differences.
To the first order in $\varepsilon_L$, the gravity modulation is out of phase
with the long-wave elevation, with the short waves experiencing a lower effective
gravity on the long-wave crests and a higher effective gravity in the troughs.
Evaluating the orbital accelerations at $z=\eta$ and the Lagrangian gravity
correction each have an $\mathcal{O}(\varepsilon_L^2)$ contribution but of
opposite signs.
The former amplifies the gravity modulation while the latter attenuates it.
The curvilinear effects are significant at the front and rear faces of the long
wave but not near the crests and troughs.
The long-wave elevation slope has a second-order effect in $\varepsilon_L$.
Naturally, the slope is exactly zero at the crests and the troughs, and its
effects are largest at the front and rear faces of the long wave, where the
covariance between the local slope and the horizontal accelerations are largest.
Evaluating the velocities and accelerations at the surface of third-order Stokes
wave causes a correction mostly at the faces of the long wave.
Finally, numerically computing the surface velocities and accelerations of a
fully nonlinear wave allows us to exclude uncertainties associated with the
truncation errors of the Stokes expansion at first and third orders in
$\varepsilon_L$.
At this steepness, the fully-nonlinear solution is marginally different from
that of the third-order Stokes wave
(see also Appendix \ref{appendix:surface_velocity_errors} for the errors
in surface velocity estimates of the linear and third-order Stokes
approximations).

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_effective_gravities.pdf}
\caption{
  Analytical solutions for the effective gravitational acceleration
  modulation by long waves, as function of the long-wave phase, for $\varepsilon_L = 0.3$.
  Black is for Eulerian gravity of a linear wave evaluated at $z=0$;
  blue is the same as black but evaluated at $z=\eta$;
  orange is the same as blue but for Lagrangian gravity;
  green is the same as orange but in a curvilinear reference frame;
  red is the same as green but for a third-order Stokes wave;
  and purple is the same as red but for a fully-nonlinear wave.
  The elevation and surface velocities for the fully-nonlinear wave are computed
  following \citet{clamond2018accurate}.
}
\label{fig:effective_gravities}
\end{figure}

It is instructive to also examine the minimum values of the
effective gravity as function of long-wave steepness $\varepsilon_L$
(Fig. \ref{fig:effective_gravities_min}), as the minima occur at or
near the long-wave crests where the short-wave modulation is largest.
Without the Lagrangian correction to the effective gravity, the gravity
modulation is overestimated; for accelerations evaluated at the surface of a
linear wave, the maximum gravity reduction is 60\% at $\varepsilon_L = 0.4$.
The Lagrangian correction attenuates the gravity modulation, making the
gravity reduction at the crests $\approx 25\%$.
As the slope effects on the effective gravity exist only away from crests and
troughs, they may be neglected in the steady solutions where modulation is
typically evaluated at the long-wave crests.
Here, the slope effect causes a difference in the minimum effective gravity
for $\varepsilon_L \gtrsim 0.3$.
The effective gravity reduction at the crests in the case of a fully nonlinear
wave is $\approx$ 10\%, 19\%, 27\%, and 40\%, 
for $\varepsilon_L$ of 0.1, 0.2, 0.3, and 0.4, respectively.
These are similar to the fully nonlinear numerical estimates of
\citet{longuet1986eulerian,longuet1987propagation}.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/fig_effective_gravities_min.pdf}
\caption{
  As Fig. \ref{fig:effective_gravities} but showing the minimum short-wave
  gravity modulation as function of long-wave steepness $\varepsilon_L$.
}
\label{fig:effective_gravities_min}
\end{figure}
